TacticToe: Learning to Prove with Tactics

### 2018-03-23 tactictoe_jv.tex
Change terminology: proof scripts -> tactic proofs

\paragraph{Resolution}
The search ends when the algorithm reaches one of these 3 conditions:
\begin{itemize}
\item It finds a proof, i.e., the root node is solved. In this case,
the search returns a minimized and
prettified proof script (see Section~\ref{sec:proofdisplay}).
\item It saturates, i.e., there are no tactics to be applied to any open goal.
This occurs less than 0.1 percent of the time in the full-scale experiment.
And this happens only at the beginning of the \holfour library, where the
training
data
has very few tactics.
\item It exceeds a given time limit (60 seconds by default).
\end{itemize}

@@

First, we develop a tactic which expects a list of premises as an
argument and calls \eprover on the goal together with the premises translated
to a first-order problem. If \eprover
succeeds on the problem, the premises used in the external proof are used to
reconstruct the proof inside \holfour with \metis. Giving a special status
to \eprover is essential as \eprover calls do not appear in human
proof scripts. Here, we use an even higher timeout (5 seconds) and
a larger number of predicted premises (128). We also try \eprover
first every time a new goal is created. Since calls to \eprover are
computationally expensive, they are run in parallel and asynchronously. To
update the MCTS values, the
result of each thread is back-propagated in the tree after completion. This
avoids slowing down \tactictoe's search loop. The number of asynchronous calls
to \eprover that can be executed at the same time is limited by the
number of cores available to the user.

@@

\subsection{Methodology}
The evaluation imitates the construction of the library: For each theorem only
the previous human proofs are known. These are used as the learning base for
the predictions.
To achieve this scenario we re-prove all theorems during a modified build of
\holfour.
As theorems are proved, their tactical proofs and their statements are
recorded and included in the training examples.
For each theorem we first attempt to run the \tactictoe search with a time
limit of 60 seconds before processing the original proof script.
In this way, the fairness of the experiments is guaranteed by construction:
only previously declared \sml values (essentially tactics, theorems and
simpsets) are accessible to \tactictoe.

@@

\begin{table}[ht]
\centering\ra{1.3}
\small
\begin{tabular}{llcc}
\toprule
 Technique & Parameters & Solved (1s) & Solved (60s)\\
\midrule
Tactic time out & 0.02s && 154\\
                & 0.05s (default) && 156\\
                & 0.1s && 154\\
\midrule
Orthogonalization & none & 105 & 156 \\
                  & $radius = 10$ & 121 & 156 \\
                  & $radius = 20$ (default) & 121 & 156 \\
                  & $radius = 40$ & 124 & 156 \\
\midrule
Abstraction       & none (default)  && 156\\
                  & $theorems = 8$  && 195\\
                  & $theorems = 16$ && 199\\
                  & $theorems = 32$ && 195\\
\midrule
MCTS policy & $c_{policy} = 0.4$ && 149\\
            & $c_{policy} = 0.5$ (default) && 156\\
            & $c_{policy} = 0.6$ && 153\\
\midrule
MCTS evaluation & none (default + best abstraction) && 199\\
				& $c_{exploration} = 1$ && 201\\
				& $c_{exploration} = 2$ && 203\\
				& $c_{exploration} = 4$ && 198\\
\midrule
ATP integration & none (default + best abstraction && 203\\
                &  + best MCTS evaluation) &&\\
                & \metis 0.1s && 216\\
                & \metis 0.2s && 212\\
                & \metis 0.4s && 212\\
                & \eprover && 213\\
                & \metis 0.1s + \eprover && 218\\
\bottomrule
\end{tabular}
\caption{\label{tab:tuning} Number of problem solved with different set of
parameters for \tactictoe on a training set of 273 theorems.}
\end{table}

@@

\paragraph{Reinforcement learning}
All our feature vectors have been learned form human proofs. We now can now
also add goal-tactic pairs that appears in the last proof of \tactictoe. To
prevent
duplication of effort, orthogonalization of those
tactics is essential to have a beneficial effect.
Since recording and re-proving are intertwined during evaluation, the
additional data is available for the next proof search.
The hope is that the algorithm will improve faster by learning from its own
discovered proofs than from the human proof
scripts~\cite{DBLP:conf/cade/Urban07}. Side experiments show that
this one shot reinforcement learning method increases \tactictoe's success rate
by less than a percent.

@@

\begin{figure}[h]
\centering
\begin{tikzpicture}[scale=1]
\begin{axis}[
  legend style={anchor=north east, at={(0.9,0.9)}},
  width=0.5*\textwidth,
  height=0.5*\textwidth,
  ymin=0, ymax=3000,
  xmin=0, xmax=10,
  xtick={},
  ytick={},
  cycle list name=my black]
\addplot table[x=length, y=proofs] {data/tactictoe_proof_length};
\addplot table[x=length, y=proofs] {data/original_proof_length};
\end{axis}
\end{tikzpicture}
\begin{tikzpicture}[scale=1]
\begin{axis}[
  legend style={anchor=north east, at={(0.9,0.9)}},
  width=0.5*\textwidth,
  height=0.5*\textwidth,
  ymin=0, ymax=120,
  xmin=10, xmax=50,
  xtick={},
  ytick={},
  cycle list name=my black]
\addplot table[x=length, y=proofs] {data/tactictoe_proof_length};
\addplot table[x=length, y=proofs] {data/original_proof_length};
\legend{\tactictoe proofs,human proofs}
\end{axis}
\end{tikzpicture}
\caption{\label{fig:distrib}Number of proof scripts with $x$ tactic units.}
\end{figure}

@@

\section{Proof Recording}\label{sec:recording}
Recording proofs in an LCF-style proof assistant can be done at different
levels.
In \holfour all existing approaches relied on modifying the kernel. This was
used
either to export the primitive inference
steps~\cite{Wong95recordingand,DBLP:conf/itp/KumarH12}
or to record dependencies of theorems~\cite{tgck-cpp15}. This was not suitable
for our
purpose of learning proving strategies at the intermediate tactic level. We
therefore
discuss recording proofs in an LCF-style proof assistant, with the focus on
\holfour
in this section.
%
Rather than relying on the underlying programming language, we parse the
original theory file containing proof scripts. This enables us to extract the
code of each tactic.
Working with the string representation of a tactic is better than working with
its value:
\begin{itemize}
\item Equality between two functions is easy to compute from
their string representation, which allows us to avoid
predicting the same tactic repeatedly.
\item
The
results of
\tactictoe can be returned in a readable form to the user. In this way, the
user can learn from the feedback, analyze the proof and possibly improve on it.
Furthermore \tactictoe does not need to be installed to run
proof scripts generated by \tactictoe. In this way, the further development of
\tactictoe does not affect the robustness of \holfour.
\item
It is difficult (probably impossible) to transfer \sml values
between theories, since they are not run in the same \holfour session. In contrast,
\sml code can be exported and imported.
\item To produce a tactic value from a code is easy in \sml, which can be
achieved by using a reference to a tactic and updating it with the function
\texttt{use}.
\end{itemize}

@@

Even when developing one theory, the context is continually changing:
modules are opened and local identifiers are defined. Therefore, it is unlikely
that code extracted from proof scripts is interpretable in any other part of
\holfour without any post-processing.
To solve this issue, we recursively replace each local identifier by its
definition until we are able to write any expression with global identifiers
only. Similarly, we prefix each global identifier by its module.
We call this process \emph{globalization}. The result is a standalone \sml code
executable in any \holfour theory.  In the absence of side effects, it is
interpreted in the same way across theories.
Therefore, we can guarantee that the behavior of recorded stateless tactics
does not change. Even with some stateful tactics, the prediction algorithm is
effective because most updates on the state of \holfour increase the
strength of stateful tactics.

\subsection{Implementation}
We describe in more detail our implementation of the recording algorithm. It
consists of 4 phases: proof script extraction, proof script globalization,
tactic unit wrapping, and creation of goal-tactic pairs.

@@

Running the globalized version of a tactic proof will have the exact same
effect as the original. But since we want to extract information from this
script in the form of tactics and their input goals, we modify it further.
In particular, we need to define at which level we should record the tactics in
the tactic proof. The simplest idea would be to record all \sml subexpressions
of type \texttt{tactic}. However, it will damage the quality of our data by
dramatically increasing the number of tactics associated with a single goal.
Imagine a proof script of the form \texttt{A THEN B THEN C}; then the tactics
\texttt{A}, \texttt{A THEN B} and \texttt{A THEN B THEN C} would be valid
advice for something close to their common input goal. The tactic
\texttt{A THEN B THEN C} is likely to be too specific. In contrast, we can consider
the tactic \texttt{REPEAT A} that is repetitively calling the tactic \texttt{A}
until
\texttt{A} has no effect. For such calls, it is often preferable to record
a single call to \texttt{REPEAT A} rather than multiple calls to \texttt{A}.
Otherwise branches after each call to \texttt{A} would be necessary as part
of proof search.
To sum up, we would like to record only the most general tactics which
make the most progress on a goal. As a trade-off between these two objectives,
we split proofs into tactic units.

@@

To produce the final recording script, we encapsulate each tactic unit
in a recording function \texttt{R} (see Example~\ref{ex:wrap}). In order to
record
tactics in all \holfour theories, we replace the original proof by the
recording script in each theory and rebuild the \holfour library.

@@

When \tactictoe finds a proof of an initial goal, it returns a search tree
$\mathfrak{T}$ where the root node containing the initial goal is solved.
In order to transform this search tree into a proof script, we need to extract
the tactics that contributed to the proof and combine them using tactic
combinators.

@@

The extracted proof script of $\mathfrak{T}$ is $P_{goal}(g_{root})$.
We minimally improve it by substituting \texttt{THENL} by \texttt{THEN} when the
list of goals is a singleton and removing \texttt{THENL []} during the
extraction phase.
Further post-processing such as
eliminating unnecessary tactics and theorems has been developed and
improves the user experience greatly~\cite{DBLP:conf/sefm/Adams15}.

\paragraph{Proof Length Minimization}
A fast and simple minimization is applied when processing the final proof. If
the tactic \texttt{A THEN B} appears in the proof and has the same effect as
\texttt{B} then we can replace \texttt{A THEN B} by \texttt{B} in the script.
Optionally, stronger minimization can be obtained
by rerunning \tactictoe with a low prior
policy coefficient, no prior evaluation and a database of tactics that contains
tactics from the discovered proof only.

@@

\paragraph{Embellishment}
Without embellishment, the returned tactic is barely readable as it contains
information to guarantees that each \sml subterm is interpreted in the way in
any context. Since we return the proof at a point where the \sml interpreter is
in a specific state, we can strip unnecessary information, such as
module prefixes. If possible, we group all local declarations in the
script under a single \texttt{let} binding at the start of the script. We
also replace extended terms by their quoted version.
All in all, if a prettified tactic $t_p$ has the same effect its original
$t_o$, we replace $t_p$ by $t_o$ in the proof.\\

The total effect of these minimization and embellishment techniques
on the readability of a proof script is depicted in
Example~\ref{ex:pretty}.

@@

\section{Conclusion}\label{sec:concl}
We proposed a new proof assistant automation technique which combines
tactic-based proof search with machine learning prediction.
Its implementation, \tactictoe, achieves an overall success rate of 66.4\%
on 7164 theorems of the \holfour standard library, surpassing \eprover
with auto-schedule. Its
effectiveness is especially visible on
theories which use inductive data structures, specialized decision procedures,
and custom built simplification sets.
Thanks to the learning abilities of \tactictoe, the generated proof scripts
often reveal the high-level structure of the proof. %For these reasons,
We therefore believe that predicting ITP tactics based on the current goal
features is a very reasonable approach to automatically guiding proof search,
and that accurate predictions can be obtained by learning from the knowledge
available in today's large formal proof corpora.

>>>>

\paragraph{Resolution}
The search ends when the algorithm reaches one of these 3 conditions:
\begin{itemize}
\item It finds a proof, i.e., the root node is solved. In this case,
the search returns a minimized and
prettified tactic proof (see Section~\ref{sec:proofdisplay}).
\item It saturates, i.e., there are no tactics to be applied to any open goal.
This occurs less than 0.1 percent of the time in the full-scale experiment.
And this happens only at the beginning of the \holfour library, where the
training
data
has very few tactics.
\item It exceeds a given time limit (60 seconds by default).
\end{itemize}

@@

First, we develop a tactic which expects a list of premises as an
argument and calls \eprover on the goal together with the premises translated
to a first-order problem. If \eprover
succeeds on the problem, the premises used in the external proof are used to
reconstruct the proof inside \holfour with \metis. Giving a special status
to \eprover is essential as \eprover calls do not appear in human-written
tactic proofs. Here, we use an even higher timeout (5 seconds) and
a larger number of predicted premises (128). We also try \eprover
first every time a new goal is created. Since calls to \eprover are
computationally expensive, they are run in parallel and asynchronously. To
update the MCTS values, the
result of each thread is back-propagated in the tree after completion. This
avoids slowing down \tactictoe's search loop. The number of asynchronous calls
to \eprover that can be executed at the same time is limited by the
number of cores available to the user.

@@

\subsection{Methodology}
The evaluation imitates the construction of the library: For each theorem only
the previous human proofs are known. These are used as the learning base for
the predictions.
To achieve this scenario we re-prove all theorems during a modified build of
\holfour.
As theorems are proved, their tactical proofs and their statements are
recorded and included in the training examples.
For each theorem we first attempt to run the \tactictoe search with a time
limit of 60 seconds before processing the original tactic proof.
In this way, the fairness of the experiments is guaranteed by construction:
only previously declared \sml values (essentially tactics, theorems and
simpsets) are accessible to \tactictoe.

@@

\begin{table}[ht]
\centering\ra{1.3}
\small
\begin{tabular}{llcc}
\toprule
 Technique & Parameters & Solved (1s) & Solved (60s)\\
\midrule
Tactic time out & 0.02s && 154\\
                & 0.05s (default) && 156\\
                & 0.1s && 154\\
\midrule
Orthogonalization & none & 105 & 156 \\
                  & $radius = 10$ & 121 & 156 \\
                  & $radius = 20$ (default) & 121 & 156 \\
                  & $radius = 40$ & 124 & 156 \\
\midrule
Abstraction       & none (default)  && 156\\
                  & $theorems = 8$  && 195\\
                  & $theorems = 16$ && 199\\
                  & $theorems = 32$ && 195\\
\midrule
  MCTS policy & $c_{\mathit{policy}} = 0.4$ && 149\\
            & $c_{\mathit{policy}} = 0.5$ (default) && 156\\
            & $c_{\mathit{policy}} = 0.6$ && 153\\
\midrule
MCTS evaluation & none (default + best abstraction) && 199\\
        & $c_{\mathit{exploration}} = 1$ && 201\\
        & $c_{\mathit{exploration}} = 2$ && 203\\
        & $c_{\mathit{exploration}} = 4$ && 198\\
\midrule
ATP integration & none (default + best abstraction && 203\\
                &  + best MCTS evaluation) &&\\
                & \metis 0.1s && 216\\
                & \metis 0.2s && 212\\
                & \metis 0.4s && 212\\
                & \eprover && 213\\
                & \metis 0.1s + \eprover && 218\\
\bottomrule
\end{tabular}
\caption{\label{tab:tuning} Number of problem solved with different set of
parameters for \tactictoe on a training set of 273 theorems.}
\end{table}

@@

\paragraph{Reinforcement learning}
All our feature vectors have been learned form human proofs. We now can now
also add goal-tactic pairs that appears in the last proof of \tactictoe. To
prevent
duplication of effort, orthogonalization of those
tactics is essential to have a beneficial effect.
Since recording and re-proving are intertwined during evaluation, the
additional data is available for the next proof search.
The hope is that the algorithm will improve faster by learning from its own
discovered proofs than from the human-written proofs~\cite{DBLP:conf/cade/Urban07}. Side experiments show that
this one shot reinforcement learning method increases \tactictoe's success rate
by less than a percent.

@@

\begin{figure}[h]
\centering
\begin{tikzpicture}[scale=1]
\begin{axis}[
  legend style={anchor=north east, at={(0.9,0.9)}},
  width=0.5*\textwidth,
  height=0.5*\textwidth,
  ymin=0, ymax=3000,
  xmin=0, xmax=10,
  xtick={},
  ytick={},
  cycle list name=my black]
\addplot table[x=length, y=proofs] {data/tactictoe_proof_length};
\addplot table[x=length, y=proofs] {data/original_proof_length};
\end{axis}
\end{tikzpicture}
\begin{tikzpicture}[scale=1]
\begin{axis}[
  legend style={anchor=north east, at={(0.9,0.9)}},
  width=0.5*\textwidth,
  height=0.5*\textwidth,
  ymin=0, ymax=120,
  xmin=10, xmax=50,
  xtick={},
  ytick={},
  cycle list name=my black]
\addplot table[x=length, y=proofs] {data/tactictoe_proof_length};
\addplot table[x=length, y=proofs] {data/original_proof_length};
\legend{\tactictoe proofs,human proofs}
\end{axis}
\end{tikzpicture}
\caption{\label{fig:distrib}Number of tactic proofs with $x$ tactic units.}
\end{figure}

@@

\section{Proof Recording}\label{sec:recording}
Recording proofs in an LCF-style proof assistant can be done at different
levels.
In \holfour all existing approaches relied on modifying the kernel. This was
used
either to export the primitive inference
steps~\cite{Wong95recordingand,DBLP:conf/itp/KumarH12}
or to record dependencies of theorems~\cite{tgck-cpp15}. This was not suitable
for our
purpose of learning proving strategies at the intermediate tactic level. We
therefore
discuss recording proofs in an LCF-style proof assistant, with the focus on
\holfour
in this section.
%
Rather than relying on the underlying programming language, we parse the
original script file containing tactic proofs. This enables us to extract the
code of each tactic.
Working with the string representation of a tactic is better than working with
its value:
\begin{itemize}
\item Equality between two functions is easy to compute from
their string representation, which allows us to avoid
predicting the same tactic repeatedly.
\item
The
results of
\tactictoe can be returned in a readable form to the user. In this way, the
user can learn from the feedback, analyze the proof and possibly improve on it.
Furthermore \tactictoe does not need to be installed to run
tactic proofs generated by \tactictoe. In this way, the further development of
\tactictoe does not affect the robustness of \holfour.
\item
It is difficult (probably impossible) to transfer \sml values
between theories, since they are not run in the same \holfour session. In contrast,
\sml code can be exported and imported.
\item To produce a tactic value from a code is easy in \sml, which can be
achieved by using a reference to a tactic and updating it with the function
\texttt{use}.
\end{itemize}

@@

Even when developing one theory, the context is continually changing:
modules are opened and local identifiers are defined. Therefore, it is unlikely
that code extracted from a tactic proof is interpretable in any other part of
\holfour without any post-processing.
To solve this issue, we recursively replace each local identifier by its
definition until we are able to write any expression with global identifiers
only. Similarly, we prefix each global identifier by its module.
We call this process \emph{globalization}. The result is a standalone \sml code
executable in any \holfour theory.  In the absence of side effects, it is
interpreted in the same way across theories.
Therefore, we can guarantee that the behavior of recorded stateless tactics
does not change. Even with some stateful tactics, the prediction algorithm is
effective because most updates on the state of \holfour increase the
strength of stateful tactics.

\subsection{Implementation}
We describe in more detail our implementation of the recording algorithm. It
consists of 4 phases: tactic proof extraction, tactic proof globalization,
tactic unit wrapping, and creation of goal-tactic pairs.

@@

Running the globalized version of a tactic proof will have the exact same
effect as the original. But since we want to extract information from this
proof in the form of tactics and their input goals, we modify it further.
In particular, we need to define at which level we should record the tactics in
the tactic proof. The simplest idea would be to record all \sml subexpressions
of type \texttt{tactic}. However, it will damage the quality of our data by
dramatically increasing the number of tactics associated with a single goal.
Imagine a tactic proof of the form \texttt{A THEN B THEN C}; then the tactics
\texttt{A}, \texttt{A THEN B} and \texttt{A THEN B THEN C} would be valid
advice for something close to their common input goal. The tactic
\texttt{A THEN B THEN C} is likely to be too specific. In contrast, we can consider
the tactic \texttt{REPEAT A} that is repetitively calling the tactic \texttt{A}
until
\texttt{A} has no effect. For such calls, it is often preferable to record
a single call to \texttt{REPEAT A} rather than multiple calls to \texttt{A}.
Otherwise branches after each call to \texttt{A} would be necessary as part
of proof search.
To sum up, we would like to record only the most general tactics which
make the most progress on a goal. As a trade-off between these two objectives,
we split proofs into tactic units.

@@

To produce the final recording tactic proof, we encapsulate each tactic unit
in a recording function \texttt{R} (see Example~\ref{ex:wrap}). In order to
record
tactics in all \holfour theories, we replace the original proof by the
recording proof in each theory and rebuild the \holfour library.

@@

When \tactictoe finds a proof of an initial goal, it returns a search tree
$\mathfrak{T}$ where the root node containing the initial goal is solved.
In order to transform this search tree into a tactic proof, we need to extract
the tactics that contributed to the proof and combine them using tactic
combinators.

@@

The extracted tactic proof of $\mathfrak{T}$ is $P_{goal}(g_{root})$.
We minimally improve it by substituting \texttt{THENL} by \texttt{THEN} when the
list of goals is a singleton and removing \texttt{THENL []} during the
extraction phase.
Further post-processing such as
eliminating unnecessary tactics and theorems has been developed and
improves the user experience greatly~\cite{DBLP:conf/sefm/Adams15}.

\paragraph{Proof Length Minimization}
A fast and simple minimization is applied when processing the final proof. If
the tactic \texttt{A THEN B} appears in the proof and has the same effect as
\texttt{B} then we can replace \texttt{A THEN B} by \texttt{B} in the proof.
Optionally, stronger minimization can be obtained
by rerunning \tactictoe with a low prior
policy coefficient, no prior evaluation and a database of tactics that contains
tactics from the discovered proof only.

@@

\paragraph{Embellishment}
Without embellishment, the returned tactic is barely readable as it contains
information to guarantees that each \sml subterm is interpreted in the way in
any context. Since we return the proof at a point where the \sml interpreter is
in a specific state, we can strip unnecessary information, such as
module prefixes. If possible, we group all local declarations in the
proof under a single \texttt{let} binding at the start of the proof. We
also replace extended terms by their quoted version.
All in all, if a prettified tactic $t_p$ has the same effect its original
$t_o$, we replace $t_p$ by $t_o$ in the proof.\\

The total effect of these minimization and embellishment techniques
on the readability of a tactic proof is depicted in
Example~\ref{ex:pretty}.

@@

\section{Conclusion}\label{sec:concl}
We proposed a new proof assistant automation technique which combines
tactic-based proof search with machine learning prediction.
Its implementation, \tactictoe, achieves an overall success rate of 66.4\%
on 7164 theorems of the \holfour standard library, surpassing \eprover
with auto-schedule. Its
effectiveness is especially visible on
theories which use inductive data structures, specialized decision procedures,
and custom built simplification sets.
Thanks to the learning abilities of \tactictoe, the generated tactic proofs
often reveal the high-level structure of the proof. %For these reasons,
We therefore believe that predicting ITP tactics based on the current goal
features is a very reasonable approach to automatically guiding proof search,
and that accurate predictions can be obtained by learning from the knowledge
available in today's large formal proof corpora.
